---
title: "Getting Started"
description: >
output: html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

## About TravelSurveyTools

The `travelSurveyTools` package provides tools for R users to aid use of data from household travel surveys. Some possible uses include creating custom cross tabs, labeling data, and calculating trip rates.

## Data Assumptions

`travelSurveyTools` assumes the the data have the structure shown below. If this does not reflect the structure of your data

### hts_data

hts_data is a list of five core tables:

#### hh

Household dataset

-   hh_id: 8 digit household ID
-   survey variables asked on a household level
-   hh_weight: household weight

#### person

Person dataset

-   hh_id: 8 digit household ID
-   person_id: 10 digit person ID
-   survey variables asked on a person level
-   person_weight: person weight

#### day

Day dataset

-   hh_id: 8 digit household ID
-   person_id: 10 digit person ID
-   day_id: 12 digit day ID
-   survey variable asked on a day level
-   day_weight: day weight

#### trip

Trip dataset

-   hh_id: 8 digit household ID
-   person_id: 10 digit person ID
-   day_id: 12 digit day ID
-   trip_id: 13 digit trip ID
-   survey variables asked on a trip level
-   trip_weight: trip weight

#### vehicle

Vehicle dataset

-   hh_id: 8 digit household ID
-   vehicle_id: 10 digit vehicle ID
-   survey responses asked on a vehicle level
-   hh_weight: household weight

### Codebook

In addition to data from the household travel survey. The codebook is also required. The codebook is assumed to be in two parts:

#### variable_list

A dataset containing information about all variables existing in the hh, person, day, trip, and vehicle tables. The variables are as follows:

-   variable: Name of the variable
-   is_checkbox: The variable is a 'Select all that Apply' question
-   hh: The variable exists in the hh table
-   person: The variable exists in the person table
-   day: The variable exists in the day table
-   trip: The variable exists in the trip table
-   vehicle: The variable exists in the vehicle table
-   location: The variable exists in the location table
-   data_type: Data type of the variable
-   description: A description of the variable
-   logic: Conditions where the variable should have a value
-   shared_name: the shared name of checkbox variable or the variable name for non-checkbox variables

#### value_labels

A dataset containing the values for all variables found in variable_list The variables are as follows:

-   variable: Name of the variable
-   value: The numeric value of the variable
-   label: What the numeric value of the variable represents
-   label_value: value concatenated with the label (e.g., 11 85 or older)
-   val_order: order for each variable label to appear in

## Using travelSurveyTools

### Prepping the Data

In order to create summaries of our data we first need to prepare our data. We can do this by using `hts_prep_data`. This will return a categorical (cat) and numeric (num) (if applicable) prepped data table that can be used to create summaries. 

```{r, hts_prep_data, eval=TRUE, echo=TRUE }

library(travelSurveyTools)
library(data.table)
library(srvyr)

# Load data
data("test_data")
data("variable_list")

DT = hts_prep_data(summarize_var = 'speed_mph',
                   variables_dt = variable_list,
                   data = test_data)

```

Numeric variables will be automatically binned in `hts_prep_data` to create categorical summaries. Here we can make a categorical summary of a numeric variable using `hts_summary`. 

```{r, hts_summary_cat, eval=TRUE, echo=TRUE}

speed_cat_summary = hts_summary(prepped_dt = DT$cat, 
                                summarize_var = 'speed_mph',
                                summarize_by = NULL,
                                summarize_vartype = 'categorical',
                                weighted = FALSE)

speed_cat_summary$summary


```

Additionally, for numeric variables we can create numeric summaries. 

```{r, hts_summary_num, eval=TRUE, echo=TRUE}

speed_num_summary = hts_summary(prepped_dt = DT$num, 
                                summarize_var = 'speed_mph',
                                summarize_by = NULL,
                                summarize_vartype = 'numeric',
                                weighted = FALSE)

speed_num_summary$summary

```


### Using Weighted Data

Additionally, we can use weighted data by setting `weighted = TRUE` and specifying the name of the weight to be used (`wtname`).


```{r, weighted_example, eval=TRUE, echo=TRUE}

speed_cat_summary = hts_summary(prepped_dt = DT$cat, 
                                summarize_var = 'speed_mph',
                                summarize_by = NULL,
                                summarize_vartype = 'categorical',
                                weighted = TRUE,
                                wtname = 'trip_weight')

speed_cat_summary$summary

```

### Calculating Standard Errors

Additionally, by specifying `se = TRUE` we can calculate standard errors.

```{r, mode_type_se_example, echo=TRUE, eval=TRUE}

speed_cat_summary = hts_summary(prepped_dt = DT$cat, 
                                summarize_var = 'speed_mph',
                                summarize_by = NULL,
                                summarize_vartype = 'categorical',
                                weighted = TRUE,
                                wtname = 'trip_weight',
                                se = TRUE)

speed_cat_summary$summary

```

### Summarizing Two Variables

If we want summarize a variable by another variable (e.g., mode type by a person's race, mode_type by a person's ethnicity) we can use the `summarize_by` argument.

```{r, mode_type_race_example, echo=TRUE, eval=TRUE}

DT = hts_prep_data(summarize_var = 'mode_type',
                   summarize_by = 'race',
                   variables_dt = variable_list,
                   data = test_data)

mode_by_race_summary = hts_summary(prepped_dt = DT$cat, 
                                   summarize_var = 'mode_type',
                                   summarize_by = 'race',
                                   summarize_vartype = 'categorical',
                                   weighted = TRUE,
                                   wtname = 'trip_weight',
                                   se = TRUE)


mode_by_race_summary$summary

```

`summarize_by` can be used with an unlimited amount of variables. To use more than one `summarize_by` variable pass a list to the argument. 

```{r, mode_type_ethnicity_example, echo=TRUE, eval=TRUE}

DT = hts_prep_data(summarize_var = 'mode_type',
                   summarize_by = c('race', 'ethnicity'),
                   variables_dt = variable_list,
                   data = list('hh' = hh,
                               'person' = person,
                               'day' = day,
                               'trip' = trip,
                               'vehicle' = vehicle))

mode_by_race_ethnicity_summary = hts_summary(prepped_dt = DT$cat, 
                                             summarize_var = 'mode_type',
                                             summarize_by = c('race', 'ethnicity'),
                                             summarize_vartype = 'categorical',
                                             weighted = TRUE,
                                             wtname = 'trip_weight',
                                             se = TRUE)


head(mode_by_race_ethnicity_summary$summary$wtd, 10)

```


### Calculating trip rates

`hts_summary` can also be used to calculate trip rates.

```{r, trip_rates_example}

DT = hts_prep_triprate(summarize_by = 'employment',
                       variables_dt = variable_list,
                       trip_name = 'trip',
                       day_name = 'day',
                       hts_data = list('hh' = hh,
                                       'person' = person,
                                       'day' = day,
                                       'trip' = trip,
                                       'vehicle' = vehicle))

trip_rate_by_employment_summary = hts_summary(prepped_dt = DT$num, 
                                              summarize_var = 'num_trips_wtd',
                                              summarize_by = 'employment',
                                              summarize_vartype = 'numeric',
                                              weighted = TRUE,
                                              wtname = 'day_weight',
                                              se = TRUE)

head(trip_rate_by_employment_summary$summary$wtd, 10)

```

### Labeling Values

To label values we can use `factorize_column`.

```{r, factorize_column}

trip_rate_by_employment_summary$summary$wtd$employment =  factorize_column(
  trip_rate_by_employment_summary$summary$wtd$employment,
  'employment',
  value_labels,
  variable_colname = 'variable',
  value_colname = 'value',
  value_label_colname = 'label',
  value_order_colname = 'val_order'
)


trip_rate_by_employment_summary$summary$wtd

```


### Creating Visuals using hts_summary output

`hts_summary` creates outputs that can easily be used to create visuals.

```{r, ggplot_example, echo=TRUE, eval=TRUE}

library(ggplot2)

p = ggplot(
  trip_rate_by_employment_summary$summary$wtd, 
  aes(x = mean, y = employment)) +
  geom_bar(stat = 'identity') + 
  geom_errorbar(
    aes(xmin = (mean - mean_se), 
        xmax = (mean + mean_se),
        width = .2)
  ) + 
  labs(x = 'Mean Trip Rate',
       y = 'Employment')  +
  scale_y_discrete(labels = function(x) stringr::str_wrap(x, width = 50),
                   limits = rev)
  
  print(p)

```
